Switching Optimisation in Huffman Code for Power Efficient Data Transmission

### 2015-02-27 Switche Paper 1st approache/elsarticle-template-num.tex
up

Genetic algorithm (GA) has widely been applied to solve a range of combinatorial problem in complex VLSI design~\cite{chan1991macro,bright1996genetic,cohoon1987genetic}. For this reason, the GA is identified as a potential solution to the above mentioned problem. GA is one the of the most popular bio-inspired meta-heuristic algorithm inspired by the natural evolution of species~\cite{gen1}. It is a population based algorithm starts with the generation of a random initial population. The population contain a set of feasible solutions called individuals, that are usually far from the optimal. The GA optimisation process uses a set of natural genetic operators such as selection, crossover and mutation to converge to the optimal solution.
\subsection{ Data preparation and population generation}
A genome is an expansion of DNA sequence which is a set of nucleotides that encodes a protein. These nucleotides are bounded together and each order define a specific protein. Three nucleotides form a triplet and interpreted as a specific amino acid ~\cite{Harvey00}. For this reason, the whole sequence is divided into a set of triplets.  As each DNA sequence contains a combination of four nucleobases\textemdash guanine (G), adenine (A), thymine (T), and cytosine (C), therefore we will have $4^3=64$ possible triplets. As mentioned earlier that the classical Huffman code requires reading the whole sequence prior to encoding. So, in this case, the whole DNA sequence is read considering different triplets as distinct symbols. Frequencies for each of the triplets are calculated and  the adjacency matrix (a square matrix) is also created which represents how many times a triplet occurs after each of the other triplets.  After that, a random set of Huffman trees are generated for the input symbols (triplets) which are considered as the initial population for the GA.
\subsection{Objective function}

The main objective of the proposed approach is to generate a set of codewords that contain the minimum number of switches in order to minimise the power consumption. The objective function of switching optimisation algorithm can be formulated as follow :

\textbf{Definition:} Let $C=\left\{C_{1},C_{2}...C_{n}\right\}$ be a set of codewords, let $F=\left\{F_{1},F_{2}...F_{n}\right\}$ be a set of frequencies of codes in the sequence, and let $M(n,n)$ be a square matrix represents the frequencies of the adjacency between codes.
\\
The Objective Function is to:

\begin{equation}
Minimise \sum_{i=1}^{n} (S(C_{i}) \times F_{i})~+~\sum_{i=1}^{n}\sum_{j=1}^{n} \mu(C_{i},C_{j}) \times M[i][j]  
\end{equation}
where\\
$\mu(C_{i},C_{j})=1$ if the head of the codeword $C_{i}$ and tail bit of the codeword $C_{j}$ are different.\\
$\mu(C_{i},C_{j})=0$ if the head of the codeword $C_{i}$ and tail bit of the codeword $C_{j}$ are equal.

\subsection{Genetic operations}
\subsubsection{Selection}
The first operator of the genetic algorithm is the selection. The main objective of the selection is to choose the part of the current  population to be a candidate for the different genetic operators in order to breed the next generation population. Many selection techniques have been proposed in the literature~\cite{bli95}. In this approach the  process of natural selection is maintained. First we generate a random pair numbers \textit{$rand$} between 2 and the population size, this number represent how many parents will be processed. After that, an unbiased random selection of \textit{$rand$}~individuals (solutions) from the population is maid. 
\subsubsection{Crossover}
The main operator of the GA is the crossover, allows to construct new solutions from the selected part of the population~\cite{osa14}. The selected solutions are ranked by fitness and crossover two by two from high fitness to low. The main objective of the crossover is to benefits from the two good solutions in order to generate a better solution. An internal node for each tree (solution) is selected (\textit{$node1$},\textit{$node2$}), these two nodes must have the same number of leaf nodes (contain the same number of codes). The crossover operation will create two new trees (solutions), the first child (second) contain the nodes that are not children of the \textit{$node1$} (\textit{$node2$}) and we replace the children of \textit{$node1$} (\textit{$node2$}) by the children of \textit{$node2$} (\textit{$node1$}) (see fig.1). 
\subsubsection{Mutation}
The mutation operator changes the positions of two leaf nodes of the generated children (result of the crossover)~\cite{muhlenbein1992genetic}; this introduce the diversity in the search process, this diversification strategy allow the algorithm do conference to the global optimum. The algorithm for this operator goes through the leaf nodes and change the position of two leaf nodes from a parent to another and selects the ones to change according to a fixed mutation rate (see fig.1).
\subsection{Population update}
The crossover operation aims to generate new solutions from the current solution to built the second generation of the population. Furthermore, the mutation provide the diversity on the search space solution by changing the position of  nodes on the same tree. After these two operations, the next step of the genetic algorithm is two to breed the new population (see fig.2). Firstly the new solutions (children) are added to the current population, after that the population is ranked by fitness. The algorithm remove the worst solutions until the initial size of the population is achieved. The algorithm genetic repeat these operators until the stopped criteria is achieved, in this algorithm the stopped criteria is when the objective function (number of switches) stop decreasing.
\begin{algorithm}[!btph]
\caption{Switches optimising Huffman codes}
\label{alg1}
\begin{algorithmic}[1]
\REQUIRE Textual representation of a Genome
\ENSURE Low switches Huffman codes
\STATE Generate triplet frequencies and adjacent matrix 
\STATE Generate the initial  population of trees
\REPEAT 
\STATE Select part of the population
\STATE Generate children candidates via crossover
\STATE Mutate children
\STATE Add children to population
\STATE Rank by fitness
\STATE Remove worst solutions until population limit
\UNTIL{The switches number stop decreasing}
%\algstore{myalg}
\end{algorithmic}
\end{algorithm}
\begin{figure}[tbph]
\begin{center}
\includegraphics[width=400pt,height=545pt]{Images/Drawing1-1.jpg}
\caption{The Crossover operator}
\end{center}
\label{Fig1}
\end{figure}
\begin{figure}[tbph]
\begin{center}
\includegraphics[width=350pt,height=170pt]{Images/Drawing3.jpg}
\caption{Population update for genetic algorithm}
\end{center}
\label{Fig2}
\end{figure}

@@

\begin{table}[h]
\renewcommand{\arraystretch}{1.1}
\small
\label{table4}
\caption{Comparison of performance among classical Huffman code, CCA, and OCCA without penalty}
%\begin{center}

\begin{tabular}{c  c c  c c  c c}
\hline
 & Huffman Algorithm & P \\\hline
\\\hline
Genome 1& 18.16 & 10.05 & 44.65 \\\hline
Genome 2& 24.66 &  15.63 & 36.61 \\\hline
Genome 3&18.46 &  10.66&  42.25\\\hline
Genome 4&25.18&16.26& 35.42\\\hline
Genome 5& 19.24& 16.08 &16.42 \\\hline
Genome 6& 19.61&14.96&23.71\\\hline
Genome 7& 22.40 &13.21&41.02\\\hline
Genome 8&23.87 & 17.70 &25.84\\\hline
Genome 9&17.66 &10.04&43.14\\\hline
Genome 10& 18.14 &9.89& 45.47 \\\hline
Genome 11&19.63&11.48&41.51 \\\hline
Genome 12& 17.59&11.86&32.57 \\\hline
Genome 13& 17.39&10.93&37.14\\\hline
Genome 14&18.53 &11.64&37.18 \\\hline
Genome 15&78.19 &50.60&35.28 \\\hline
Genome 16& 168.98& 102.13&39.56 \\\hline
Genome 17&217.68 &130.93 & 39.85 \\\hline
Genome 18& 243.70& 149.80&38.53 \\\hline
%\bottomrule
\end{tabular}
%\end{center}
\end{table}
Table 2 presents thee results obtained by the proposed approach and classical Huffman code. The results are presented with the total number of switches and the improvement rate. The results shows that the proposed approach optimise the power consumption of the whole genome by reducing the total number of inter- and intra-switches. In the best case the proposed approach improves power consumption over classical Huffman code by $x$, in the worst case by $x$, and on an average by $x$. The proposed approach optimise the total number of switches by swapping finding the best arrangement of nodes on the codewords tree with the best allocation of these codewords to the different frequencies (triplet). The approach continue swapping between the population individuals until a balance is found between inter-switches and intra-switches.

>>>>

Genetic algorithm (GA) has widely been applied to solve a range of combinatorial problem in complex VLSI design~\cite{chan1991macro,bright1996genetic,cohoon1987genetic}. For this reason, the GA is identified as a potential solution to the above mentioned problem. GA is one the of the most popular bio-inspired meta-heuristic algorithm inspired by the natural evolution of species~\cite{gen1}. It is a population based algorithm starts with the generation of a random initial population. The population contain a set of feasible solutions called individuals, that are usually far from the optimal. The GA optimisation process uses a set of natural genetic operators such as selection, crossover and mutation to converge to the optimal solution. The whole process is shown in algorithm 2.
\begin{algorithm}[!btph]
\caption{Switches optimising Huffman codes}
\label{alg1}
\begin{algorithmic}[1]
\REQUIRE Textual representation of a Genome
\ENSURE Low switches Huffman codes
\STATE Generate triplet frequencies and adjacent matrix 
\STATE Generate the initial  population of trees
\REPEAT 
\STATE Select part of the population
\STATE Generate children candidates via crossover
\STATE Mutate children
\STATE Add children to population
\STATE Rank by fitness
\STATE Remove worst solutions until population limit
\UNTIL{The switches number stop decreasing}
%\algstore{myalg}
\end{algorithmic}
\end{algorithm}
\subsection{ Data preparation and population generation}
A genome is an expansion of DNA sequence which is a set of nucleotides that encodes a protein. These nucleotides are bounded together and each order define a specific protein. Three nucleotides form a triplet and interpreted as a specific amino acid ~\cite{Harvey00}. For this reason, the whole sequence is divided into a set of triplets.  As each DNA sequence contains a combination of four nucleobases\textemdash guanine (G), adenine (A), thymine (T), and cytosine (C), therefore we will have $4^3=64$ possible triplets. As mentioned earlier that the classical Huffman code requires reading the whole sequence prior to encoding. So, in this case, the whole DNA sequence is read considering different triplets as distinct symbols. Frequencies for each of the triplets are calculated and  the adjacency matrix (a square matrix) is also created which represents how many times a triplet occurs after each of the other triplets.  After that, a random set of Huffman trees are generated for the input symbols (triplets) which are considered as the initial population for the GA.
\subsection{Objective function}

The main objective of the proposed approach is to generate a set of codewords for the input symbols to minimise the overall amount of switching. The objective function of the switching optimisation algorithm can be formulated as follow :

\textbf{Definition:} Let $F=\left\{F_{1},F_{2},...,F_{n}\right\}$ and $C=\left\{C_{1},C_{2},...,C_{n}\right\}$ be the set of frequencies and codewords for the input symbols $S=\left\{S_{1},S_{2},...,S_{n}\right\}$ respectively. And $M(n,n)$ be the adjacency matrix where each entry $M[i][j]$ represents how many times the input symbol $S_j$ appears after the symbol $S_i$ in the whole sequence.
\\
The Objective Function is to:

\begin{equation}
Minimise~\Big( \sum_{i=1}^{n}~(~Sw(C_{i}) \times F_{i}~)~+~\sum_{i=1}^{n}\sum_{j=1}^{n}~(~\mu(C_{i},C_{j}) \times M[i][j]~)~\Big)  
\end{equation}
where,\\\\
$Sw(C_{i})$ is the number of intra-switches in the codewords $C_{i}$\\
$\mu(C_{i},C_{j})=1$ if the head (starting) bit of the codeword $C_{j}$ and the tail (ending) bit of the codeword $C_{i}$ are different.\\
$\mu(C_{i},C_{j})=0$ if the head bit of the codeword $C_{j}$ and the tail bit of the codeword $C_{i}$ are equal.

\subsection{Genetic operations}
\subsubsection{Selection}
The first operator of the genetic algorithm is the selection. The main objective of the selection is to choose the part of the current  population to be a candidate for the different genetic operators in order to breed the next generation population. Many selection techniques have been proposed in the literature~\cite{bli95}. In this approach the  process of natural selection is maintained. At first, we generate a random even number \textit{$rand$} between 2 and the population size. This number represents how many parents will be processed. After that, an unbiased random selection of \textit{$rand$}~individuals (solutions) from the population is made. 
\subsubsection{Crossover}
The main operator of the GA is the crossover, allows to construct new solutions from the selected part of the population~\cite{osa14}. The selected solutions are ranked by fitness and sorted in descending order. After that, at a time, a pair of solution is taken from the sorted population and then crossed to generate two new solutions. This process continues until all the pairs of solutions are crossed. The main objective of the crossover is to benefits from the two solutions  to generate better solutions.
Two parent trees are crossed to find two new child trees in the following way:\\\\
\textbf{Step 1: }Select an internal node (a sub-tree) from each parent tree with equal number of nodes (the dashed selection in Fig. 1(a) and (b)). \\
\textbf{Step 2:} Put the selected sub-tree of the first tree in the place of the selected sub-tree of the second tree and vice versa (see Fig. 1(c) and (d)).\\

However, in each of the trees some symbols are repeated and some are missed, but number of repeated and missed symbols are equal. This is because we replace a sub-tree by other sub-tree without any modification. As a result each child tree contains a part from its parent tree and a part from the other tree. If the remaining part from the parent tree and the newly added part has one or more  symbols in common then these symbols will be repeated in the tree, on the other hand, the uncommon symbols between the newly added sub-tree and  the replaced sub-tree will be missed.  We solved this conflict by putting the missed symbols one by one in the place of the repeated symbol in the remaining part from the parent tree. That means, the newly added sub-tree is kept unchanged and all the replacements are done on the old part of the tree.

\subsubsection{Mutation}
The mutation operator changes the positions of two leaf nodes of the generated children (result of the crossover)~\cite{muhlenbein1992genetic}. It introduces the diversity in the search process, this diversification strategy allow the algorithm to converge to the optimal solution. The algorithm for this operator goes through the leaf nodes and change the positions of two leaf nodes from their current positions to another positions in the tree (see Fig.1(e) and (f)). The selection of the nodes to be mutated is done following a fixed mutation rate.

\subsection{Population update}
The crossover operation generates new solutions from the current solutions to build the second generation of the population. Furthermore, the mutation process provides the diversity in the search on the solution space by changing the position of the nodes in the same tree. After these two operations, the next step of the genetic algorithm is to breed the new population (see fig.2). Firstly, the new solutions (children) are added to the current population and ranked by fitness. The algorithm remove the worst solutions until the initial size of the population is achieved. The genetic algorithm  continuously repeats the selection, crossover, mutation, and population update operation until the stopping criteria is achieved. In the proposed approach the stopping criteria is when the objective function (number of switches) stop decreasing.
\begin{figure}[tbph]
\begin{center}
\includegraphics[width=400pt,height=545pt]{Images/Drawing1-1.jpg}
\caption{The Crossover operator}
\end{center}
\label{Fig1}
\end{figure}
\begin{figure}[tbph]
\begin{center}
\includegraphics[scale=0.7]{Images/Drawing_1.jpg}
\caption{Population update for genetic algorithm}
\end{center}
\label{Fig2}
\end{figure}

@@

\begin{table}[h]
\renewcommand{\arraystretch}{1.1}
\small
\label{table4}
\caption{Comparison of the total number of switches (in millions) between the Huffman code and the proposed approach}
%\begin{center}

\begin{tabular}{c  c c  c c  c c}
\hline
 Datasets & \begin{tabular}{@{}c@{}}Switches in the\\ Huffman Code\end{tabular} & \begin{tabular}{@{}c@{}}Switches in the\\ proposed approach\end{tabular}& Improvement (\%) \\\hline
Genome 1& 18.16 & 10.05 & 44.65 \\\hline
Genome 2& 24.66 &  15.63 & 36.61 \\\hline
Genome 3&18.46 &  10.66&  42.25\\\hline
Genome 4&25.18&16.26& 35.42\\\hline
Genome 5& 19.24& 16.08 &16.42 \\\hline
Genome 6& 19.61&14.96&23.71\\\hline
Genome 7& 22.40 &13.21&41.02\\\hline
Genome 8&23.87 & 17.70 &25.84\\\hline
Genome 9&17.66 &10.04&43.14\\\hline
Genome 10& 18.14 &9.89& 45.47 \\\hline
Genome 11&19.63&11.48&41.51 \\\hline
Genome 12& 17.59&11.86&32.57 \\\hline
Genome 13& 17.39&10.93&37.14\\\hline
Genome 14&18.53 &11.64&37.18 \\\hline
Genome 15&78.19 &50.60&35.28 \\\hline
Genome 16& 168.98& 102.13&39.56 \\\hline
Genome 17&217.68 &130.93 & 39.85 \\\hline
Genome 18& 243.70& 149.80&38.53 \\\hline
%\bottomrule
\end{tabular}
%\end{center}
\end{table}
Table 2 presents thee results obtained by the proposed approach and classical Huffman code. The results are presented with the total number of switches and the improvement rate. The results shows that the proposed approach optimise the power consumption of the whole genome by reducing the total number of inter- and intra-switches. In the best case the proposed approach improves power consumption over classical Huffman code by $x$, in the worst case by $x$, and on an average by $x$. The proposed approach optimise the total number of switches by swapping finding the best arrangement of nodes on the codewords tree with the best allocation of these codewords to the different frequencies (triplet). The approach continue swapping between the population individuals until a balance is found between inter-switches and intra-switches.
